---
title: "COVID-19 Dashboard"
author: "Nathaniel Schmucker" 
date: "Updated: `r format(Sys.time(), '%B %d, %Y')`"
runtime: shiny
output: 
  html_document:
    theme: paper  
---

```{r setup, include = FALSE}
library(readr)
library(dplyr)
library(stringr)
library(zoo)     # masks as.Date

# library(tigris) # only needed if refreshing shapefiles
library(sf)

library(shiny)
library(shinyWidgets)
library(htmltools)
library(leaflet)
library(highcharter)
library(DT)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

hcoptslang <- getOption("highcharter.lang")
hcoptslang$thousandsSep <- ","
options(highcharter.lang = hcoptslang)

source("./R/Utils.R")
source("./R/Functions.R")

#TODO: Update ReadMe
#TODO: Make repo public

```

```{r get-data, include = FALSE}
# Get census data from census.gov and COVID-19 data from NYTimes Github
url_census   <- "https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/totals/co-est2019-alldata.csv"
url_states   <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"

raw_census   <- readr::read_csv(url(url_census))
raw_states   <- readr::read_csv(url(url_states))
raw_counties <- readr::read_csv(url(url_counties))

# # Import shapefiles for mapping (only need to load and clean once)
# # Get data on states (simplefeatures object)
# states <- tigris::states(
#   cb = TRUE, 
#   resolution = "500k", 
#   year = 2018,
#   class = "sf"
# ) 
# 
# # Get data on counties (simplefeatures object)
# counties <- tigris::counties(
#   state = NULL,
#   cb = TRUE,
#   resolution = "500k",
#   year = 2018,
#   class = "sf"
# ) 

# Load pre-saved state and county shapefiles
states   <- sf::read_sf("./Data/states_clean.shp")
counties <- sf::read_sf("./Data/counties_clean.shp")

```

```{r clean-and-merge-data, include = FALSE}
covid_19 <-
  consolidate_levels(raw_states, raw_counties) %>% 
  add_population(raw_census) %>% 
  add_variables() %>% 
  beautify()

# Ensure combined, transformed data hasn't added/dropped rows
stopifnot(
  nrow(covid_19 %>% 
    dplyr::filter(GEOID != "36NYC")) ==
    nrow(raw_states %>%
           dplyr::filter(!state %in% c(
             "Alaska", "American Samoa", "Guam", "Hawaii", 
             "Northern Mariana Islands", "Puerto Rico", "Virgin Islands"
           )) %>% 
           dplyr::filter(!is.na(fips))
    ) +
    nrow(raw_counties %>%
           dplyr::filter(!state %in% c(
             "Alaska", "American Samoa", "Guam", "Hawaii", 
             "Northern Mariana Islands", "Puerto Rico", "Virgin Islands"
           )) %>% 
           dplyr::filter(!is.na(fips))
    )
)

# readr::write_csv(covid_19, paste0("./Data/COVID_19_", Sys.Date(), ".csv"))

# states <- states %>% 
#   sf::st_transform(crs = "+proj=longlat +datum=WGS84") %>% 
#   dplyr::filter(
#     !STATEFP %in% c(
#       "02", #Alaska
#       "60", #American Samoa
#       "69", #Commonwealth of the Northern Mariana Islands
#       "66", #Guam
#       "15", #Hawaii
#       "72", #Puerto Rico
#       "78"  #United States Virgin Islands
#     )
#   ) %>% 
#   dplyr::select(-STATENS, -AFFGEOID, -STUSPS, -NAME, -LSAD, -ALAND, -AWATER)
# 
# counties <- counties %>%
#   sf::st_transform(crs = "+proj=longlat +datum=WGS84") %>% 
#   dplyr::filter(
#     !STATEFP %in% c(
#       "02", #Alaska
#       "60", #American Samoa
#       "69", #Commonwealth of the Northern Mariana Islands
#       "66", #Guam
#       "15", #Hawaii
#       "72", #Puerto Rico
#       "78"  #United States Virgin Islands
#     )
#   ) %>% 
#   dplyr::select(-COUNTYNS, -AFFGEOID, -NAME, -LSAD, -ALAND, -AWATER)

# sf::write_sf(states, "./Data/states_clean.shp")
# sf::write_sf(counties, "./Data/counties_clean.shp")

```

### Introduction

The scale of the COVID-19 pandemic in the United States has grown rapidly since the first reported cases in Washington State in late January. As of mid-April, the U.S. has over 555,000 confirmed cases in over 1,500 counties, and over 22,000 confirmed deaths--more deaths than any other country in the world. This dashboard was created to help visualize the spread of the virus throughout the country. For more information about the sARS-CoV-2 virus and COVID-19 disease, the [CDC website](https://www.cdc.gov/coronavirus/2019-ncov/index.html) publishes up-to-date scientific information, guidance and data reports on the U.S. outbreak. 

#### A note on data
This document contains maps, charts, and tables that summarize COVID-19 cases and fatalities in the United States. The New York Times aggregates data based on reports from state and local health agencies, and it has made the data public in a [GitHub repository](https://github.com/nytimes/covid-19-data). The data is for the United States only, reported at the state and county level. For reporting purposes, all New York City counties have been consolidated to one reporting unit. Other exceptions are documents on the New York Times' GitHub page. This dashboard excludes Alaska, Hawaii, and U.S. Territories.

The COVID-19 data begins with the first reported coronavirus case in Washington State on Jan. 21, 2020. The New York Times updates its data regularly (as of mid-April, the update is daily around noon), and this dashboard retrieves the latest New York Times data each time it is opened. 
Population data comes from the [U.S. Census Bureau's](https://www.census.gov/data/tables/time-series/demo/popest/2010s-counties-total.html) 2019 estimate of resident population for counties.

#### Definitions

* **Total cases:** Number of confirmed COVID-19 cases
* **Total deaths:** Number of confirmed deaths due to COVID-19
* **Total cases per 100k:** Number of confirmed COVID-19 cases per 100k residents
* **Total deaths per 100k:** Number of confirmed deaths due to COVID-19 per 100k residents
* **New cases:** Number of newly-confirmed COVID-19 cases
* **New deaths:** Number of newly-reported deaths due to COVID-19
* **Average new cases:** 7-day rolling average number of newly-confirmed COVID-19 cases (Null if fewer than 7 days)
* **Average new deaths:** 7-day rolling average number of newly-confirmed deaths due to COVID-19 (Null if fewer than 7 days)
* **Average infection rate:** Average new cases as a percent of total cases
* **Average death rate:** Average new deaths as a percent of total cases
* **Case fatality rate:** Total deaths as a percent of total cases

*Notes:*

1. Data is subject to reporting error, as testing can yield false positives and false negatives. Many true cases are also unreported due to lack of widespread testing.
2. As the rate testing increases, the number of reported cases will increase, though the actual prevalence of disease may not be increasing at the same rate.
3. Data do not report recovered cases, and as of publication, it is not known how long patients remain contagious after they have been infected. Hence, average infection rate and average death rate will decrease as their denominators become overstated.

### Metric and data filters

```{r reactive-inputs}
shinyWidgets::selectizeGroupUI(
  id = "geoFilters",
  params = list(
    LEVEL = list(inputId = "LEVEL", title = "Reporting level:"),
    STATE_NAME = list(inputId = "STATE_NAME", title = "State:"),
    COUNTY_NAME = list(inputId = "COUNTY_NAME", title = "County:")
  ),
  label = "Filters",
  btn_label = "Clear filters",
  inline = TRUE
)

shiny::varSelectInput(
  inputId = "metric", label = "Metric:",
  data = dplyr::select(covid_19, 8:18)
)

shiny::radioButtons(
  inputId = "logAxis", label = "Log axis?",
  choices = c("No", "Yes"), selected = "No", 
  inline = TRUE
)

```

```{r reactive-filter}
filter_covid_19 <- shiny::callModule(
  module = shinyWidgets::selectizeGroupServer,
  id = "geoFilters",
  data = covid_19,
  vars = c("LEVEL", "STATE_NAME", "COUNTY_NAME")
)

```

### Current metric values

#### Map

```{r map}
renderLeaflet({
  data_frame <- filter_covid_19() %>%
    prep_output_data(!!input$metric, log_axis = "No", latest = TRUE) %>% 
    add_nyc_counties()
  
  spatial_data <- if(data_frame$LEVEL[1] == "State") {states} else {counties}
  
  map <- tigris::geo_join(
    spatial_data = spatial_data,
    data_frame = data_frame,
    by_sp = "GEOID",
    by_df = "GEOID",
    how = "inner"
  )

  pal <- leaflet::colorNumeric(
    palette = "YlOrRd",
    domain = map$VALUE_FULL
  )
  
  leaflet(map) %>%
    leaflet::addPolygons(
      color = "#444444", weight = 1, smoothFactor = 0.5,
      opacity = 1.0, fillOpacity = 0.5,
      fillColor = ~pal(VALUE_FULL),
      label = ~lapply(LABEL, htmltools::HTML),
      highlightOptions = leaflet::highlightOptions(
        color = "white", weight = 2, bringToFront = TRUE)
    )
})

```

#### Chart

```{r chart-latest}
highcharter::renderHighchart({
  graph_bars(filter_covid_19(), !!input$metric)
})

```

#### Table

```{r data-table-latest}
DT::renderDataTable({
  filter_covid_19() %>% 
    prep_output_data(!!input$metric, log_axis = "No", latest = TRUE) %>% 
    dplyr::select(DATE, VALUE, STATE_NAME, COUNTY_NAME)
})

```

### Development of metric over time

#### Chart

```{r chart-all}
highcharter::renderHighchart({
  graph_timeseries(filter_covid_19(), !!input$metric, log_axis = input$logAxis)
})

```

#### Table

```{r data-table-all}
DT::renderDataTable({
  filter_covid_19() %>% 
    prep_output_data(!!input$metric, log_axis = "No", latest = FALSE) %>% 
    dplyr::select(DATE, VALUE, STATE_NAME, COUNTY_NAME) %>% 
    DT::datatable()
})

```
